Model Finding for Recursive Functions in SMT

### 2015-04-30 tex/main.tex
More text, contributions.

\ajr{(1) Encoding, (2) Evaluation, (3) Front-end support for define-funs-rec in \cvc}

@@

It is important to note that the translation $\conv$ does not preserve the models of $\varphi$.
One model $\M$ for formula~(\ref{eq:ex-after}) in the previous example interprets 
$\fargsort{s}$ as a finite set of elements $\{ u_0, \ldots, u_{14} \}$,
$\farg{s}$ as a finite map mapping containing $u_i \mapsto i$ for $i = 0, \ldots, 14$,
$c$ as $14$, 
and $s$ as the (almost constant) function:
\begin{eqnarray}
\lambda x\typ{Int}. \lite( x \teq 0, 0, \lite( x \teq 1, 1, \lite( x \teq 2, 3, \ldots, \lite( x \teq 13, 91, 105 ) \ldots )
\end{eqnarray}
In other words, $s$ is a function mapping $x$ to the sum of all positive integers between $0$ and $x$ when $0 \leq x \leq 13$,
and $105$ otherwise.
Notice that $\M$ is \emph{not} a model for the original formula~(\ref{eq:ex-before}),
since $\M$ interprets e.g. $s( -1 )$ and $s( 15 )$ as $105$.
However, under the assumption that function definitions in $\sfundefs{\Sigma}$ are logically consistent, 
we claim that $\conv(\varphi)$ is equisatisfiable to $\varphi$, for all $\varphi$.
The intuition is that the interpretation of a term such as $s( -1 )$ in the previous example is not relevant to the satisfiability of formula~(\ref{eq:ex-before}).
\ajr{add invariants, proofs}

@@

We have seen a satisfiability-preserving translation $\conv$ that significantly improves the effectiveness of known SMT techniques
for finding counterexamples for formulas involving recursive function definitions.
For convienience, the translation $\conv$ has been implemented in the latest development version of \cvc~\cite{ReyEtAl-1-RR-13}.
Function definitions $\forallf{f} \vec x. \varphi$ can be written in using the $\definefunreccmd$ command, 
proposed in the draft of SMT version 2.5~\cite{}.
For instance, the input from Example~\ref{ex:translation} can be written as:

>>>>

%\ajr{(1) Encoding, (2) Evaluation, (3) Front-end support for define-funs-rec in \cvc}

In this paper, we make the following contributions:
\begin{itemize}
\item[-] We define a satisfiability-preserving translation method for formulas containing recursive function definitions
that produces formulas for which existing model finding techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.
\item[-] We provide experimental evidence that this translation improves the effectiveness of SMT solvers \cvc and \ziii for finding counterexamples to verification conditions.
\item[-] We detail front-end support for recursive function definitions and this translation in the SMT solver \cvc.
\end{itemize}

@@

It is important to note that the translation $\conv$ does not preserve the models of $\varphi$.
One model $\M$ for formula~(\ref{eq:ex-after}) in the previous example interprets 
$\fargsort{s}$ as a finite set of elements $\{ u_0, \ldots, u_{14} \}$,
$\farg{s}$ as a finite map mapping containing $u_i \mapsto i$ for $i = 0, \ldots, 14$,
$c$ as $14$, 
and $s$ as the (almost constant) function:
\begin{eqnarray}
\lambda x\typ{Int}. \lite( x \teq 0, 0, \lite( x \teq 1, 1, \lite( x \teq 2, 3, \ldots, \lite( x \teq 13, 91, 105 ) \ldots )
\end{eqnarray}
In other words, $s$ is a function mapping $x$ to the sum of all positive integers between $0$ and $x$ when $0 \leq x \leq 13$,
and $105$ otherwise.
Notice that $\M$ is \emph{not} a model for the original formula~(\ref{eq:ex-before}),
since $\M$ interprets e.g. $s( -1 )$ and $s( 15 )$ as $105$.
However, under the assumption that function definitions in $\sfundefs{\Sigma}$ are logically consistent, 
we claim that $\conv(\varphi)$ is equisatisfiable to $\varphi$ for any input $\varphi$.
The intuition is that the interpretation of a term such as $s( -1 )$ in the previous example is not relevant to the satisfiability of formula~(\ref{eq:ex-before}).
Thus, a satisfiable (resp. unsatisfiable) response from an SMT solver on input $\conv( \varphi )$ implies the existence (resp. non-existence) of a model for $\varphi$.
Moreover, information in models for $\conv( \varphi )$ contains relevant information regarding the models of $\varphi$.
For instance, the model $\M$ for formula~(\ref{eq:ex-after}) described above interprets $c$ as $14$.
In fact, we claim that there exists a model of $\varphi$ that coincides with $\M$ on its interpretation of all functions in $\sfuns{\Sigma}$ 
other than those in $\sfundefs{\Sigma}$, i.e. $\{ s \}$ in the previous example.
From a practical point of view, this is in fact a non-issue, as $s$ is the very function that was explicitly defined by the user,
and hence already has an intended interpretation.

\ajr{add invariants, proofs}

@@

We have seen a satisfiability-preserving translation $\conv$ that significantly improves the effectiveness of known SMT techniques
for finding counterexamples for formulas involving recursive function definitions.
For convienience, the translation $\conv$ has been implemented in the latest development version of \cvc~\cite{ReyEtAl-1-RR-13} (version 1.5 pre-release).
Function definitions $\forallf{f} \vec x. \varphi$ can be written in using the $\definefunreccmd$ command, 
proposed in the draft of SMT version 2.5~\cite{}.
For instance, the input from Example~\ref{ex:translation} can be written as:
